---
title: "Week 11: Categorical Data II"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: November 18, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

# Categorical Data as Explanatory Variables

## Categorical Data as Explanatory Variable

- We have already learned how to deal with the case when a binary variable like gender is used as a explanatory (independent) variable
- In this case, gender is also a categorical variable, with two possible values
  - In recent years, large national surveys now include more categories in asking gender \pause
- We include gender in regression by transforming it into a 0-1 binary, where 1 represents woman, and 0 man (or 1 as man and 0 as woman)
- $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + e_i$
- How do intepret $\hat{\beta}_2$ (suppose this is a positive number)? \pause
- Conditioning on the same level of education, women on average have $\hat{\beta}_2$-unit higher levels abortion attitudes than men do

---

## Categorical Data as Explanatory Variable

- $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + e_i$
- Men do not explicitly appear in regression; instead, it appears as the **reference group** for women to be compared with \pause
- Including a categorical variable that has multiple variables (e.g., race, region, marital status) is analogous
- For example, if we want to add race (5 categories: White, Black, Hispanic, Asian, Others) in the above equation
- It is incorrect to directly specify $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + \hat{\beta}_3race_i + e_i$
- As race is not ordered and additive

---

## Categorical Data as Explanatory Variable

- Instead, we transform race into 5 "dummy variables" (dummy means a binary variable created from categorical variables), `white`, `black`, `hispanic`, `asian`, `others`
- If the respondent is a black, $black = 1$ and other other dummy variables $=0$
- If the respondent is a hispanic, $hispanic = 1$ and other other dummy variables $=0$ \pause
- We put all these dummy variables but one group in regression; the group that is omitted serves as the **reference** group
- $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + \hat{\beta}_3black_i +  \hat{\beta}_4hispanic_i + \hat{\beta}_5asian_i + \hat{\beta}_6others_i + e_i$ \pause
- Which group to omit is up to you, but it is related to the interpretation of the results

---

## Categorical Data as Explanatory Variable

- $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + \hat{\beta}_3black_i +  \hat{\beta}_4hispanic_i + \hat{\beta}_5asian_i + \hat{\beta}_6others_i + e_i$
- How do we interpret $\hat{\beta}_3$ (suppose this is a positive number)? \pause
- Conditioning on the same level of education and gender, blacks on average have $\hat{\beta}_3$-unit higher levels of abortion attitudes than whites do
- Whites do not explicitly appear in regression; instead, it appears as the **reference group** for other racial groups to be compared with

---

## Practice

- We are now interested in the association between marital status and religious identification, controlling for gender and education
- There are five categories of marital status, Married, Widowed, Divorced, Separated, Never married 

---

## Practice

- $Relig_i = 2.56 + 0.013edu_i + (-0.094)women_i + (-0.355)widowed_i + 0.139divorced_i + 0.376separated + 0.635nevmar_i + e_i$

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE,results="asis"}
library(stargazer)

## set your working directory - you should set your own unique one!
setwd("~/Dropbox/Teaching/SOCUA-302/Week 11")

## read csv data - this is 2021 GSS data
gss <- read.csv("GSS_SOCUA_W11.csv")
gss <- gss %>%
  mutate(relig=ifelse(relig==4,0,1))

## specify model
model4 <- lm(relig~educ+sex+factor(marital),gss)
stargazer(model4, out = "text",
          header=FALSE,
          title = "The association between education, sex, marital status and religion, GSS 2021",
          digits = 3,
          single.row = T,
          covariate.labels = c("Education","Women","Widowed","Divorced","Separated","Never married","Intercept"),
          dep.var.labels  = "Religious identification",
          omit.stat = c("rsq","f","ser"))
```


# Data Memo and R Operations

## Logistics

- Research Memo: 5-page memo presenting an interesting statistical result and discussing its social science implications
- Due **16 December**

---

## General Expectations

- Not required to go very advanced and substantively innovative (but encouraged!), but we generally expect to see 
- (1) An introduction with some literature review
  - A clear-defined topic (e.g., what is the association between education and gender ideologies/attitudes)
- (2) A brief description of the data you intend to use  
  - For example, if you want to use GSS, you may describe which year of data you use, how many individuals are included, what is their gender and race composition, etc.
  
---

## General Expectations

- (3) Clear-stated dependent (e.g., gender ideologies) and independent (years of education) variables
- (4) Start with a two-way cross-table
- (5) Proceed to a bivariate regression
- (6) Adding a set of other controlling variables in a multivariate regression setting, such as gender, age, and region
  - Try to test for interactions. For example, does the effect of years of education in gender ideologies depend on which gender the respondent is?
  - Or, is the variable a mediator, a moderator, or a multiple cause?
- (7) Visualize the relationship, possibly after step (3) and step (6)  
- (8) Conclusion and implications

---

## Logistics

- It is almost certain that you will encounter difficulties in cleaning the data
- Come to my office hour and I can help!

## Data Sources

- [General Social Survey](https://gssdataexplorer.norc.org/home) 
- [Current Population Survey](https://cps.ipums.org/cps/index.shtml) 
- [American Community Survey](https://usa.ipums.org/usa/)
- [American Decennial Census](https://usa.ipums.org/usa/) 
- [American Time Use Survey](https://timeuse.ipums.org/) 
- [National Longitudinal Survey of Youth 1979 and 1997](https://www.nlsinfo.org/investigator/pages/login)

---

## Read Data

```{r,warning=FALSE}
## set your working directory - you should set your own unique one!
setwd("~/Dropbox/Teaching/SOCUA-302/Week 11")

## read csv data - this is 2021 GSS data
gss <- read.csv("GSS_SOCUA_W11.csv")
```

---

## Cross Table

- The frequency of men and women who have and have no religious identification

\footnotesize

```{r}
## recode religion
gss <- gss %>%
  mutate(relig=ifelse(relig==4,0,1))

## cross table
source("http://pcwww.liv.ac.uk/~william/R/crosstab.r")
crosstab(gss, row.vars = "sex", 
         col.vars = "relig", 
         type = "f") ## "f" represents frequency
```

---

## Cross Table

- The proportion of people who have and have no religious identification among men and women

```{r}
## cross table
crosstab(gss, row.vars = "sex", 
         col.vars = "relig", 
         type = "r") ## "r" represents row-wise proportion 
```

---

## Cross Table

- The `crosstab` function by default returns proportion by row
- We can also change it to column-wise proportion
- The proportion of men and women among the people who have and have no religious identification

```{r}
## cross table
crosstab(gss, row.vars = "sex", 
         col.vars = "relig", 
         type = "c") ## "c" represents column-wise proportion
```

---

## Cross Table

\footnotesize 

- We can also include multiple categories as rows

```{r}
## cross table
crosstab(gss, row.vars = c("marital","sex"), 
         col.vars = "relig", 
         type = "r")
```

- Women who are widowed have the highest proportion of religious believers among all categories.

---

## Cross Table

- The proportion can also be joint percentages

```{r,eval=FALSE}
## cross table
crosstab(gss, row.vars = c("marital","sex"), 
         col.vars = "relig", 
         type = "j")
```

---

\footnotesize 

```{r,echo=FALSE}
## cross table
crosstab(gss, row.vars = c("marital","sex"), 
         col.vars = "relig", 
         type = "j")
```

---

- Although the function `crosstab` is user-written, it is pretty flexible
- Take a look at the original instructions [here](http://rstudio-pubs-static.s3.amazonaws.com/6975_c4943349b6174f448104a5513fed59a9.html) if you want more ways of cross-tabulation to explore the data

---

## Regression

- The basic format of regression in `R` is `lm(y ~ x1 + x2 + ... xn, data)`
- Save the regression model by some name, e.g., `model1 <- lm(y ~ x1 + x2 + ... xn, data)`
- You can check the regression output by `summary(model1)` \pause
- There is a more elegant way to present results by calling `stargazer()`

\footnotesize 

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE, results="asis"}
library(stargazer)
model1 <- lm(relig~educ,gss)
stargazer(model1, type = "text",
          header=FALSE,
          title = "The association between education and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression

\footnotesize 

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center",echo=FALSE, warning=FALSE, results="asis"}
library(stargazer)
model1 <- lm(relig~educ,gss)
stargazer(model1, out = "text",
          header=FALSE,
          title = "The association between education and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression

- There is a more elegant way to present results by calling `stargazer()`
- `stargazer()` can present multiple results of regressions

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE, results="asis"}
model1 <- lm(relig~educ,gss)
model2 <- lm(relig~educ+sex,gss)
stargazer(model1,model2, type = "text",
          header=FALSE,
          title = "The association between education, sex and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression

\footnotesize 

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE, results="asis"}
model1 <- lm(relig~educ,gss)
model2 <- lm(relig~educ+sex,gss)
stargazer(model1,model2, out = "text",
          header=FALSE,
          title = "The association between education, sex and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression

- We can also include an interaction term to examine whether the association between education and religious identification depends on gender \pause
- In `R`, the product of two terms is controlled by the `*` sign

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE, results="asis"}
model1 <- lm(relig~educ,gss)
model2 <- lm(relig~educ+sex,gss)
model3 <- lm(relig~educ+sex+educ*sex,gss) 
stargazer(model1,model2,model3, type = "text",
          header=FALSE,
          title = "The association between education, sex and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression 

\scriptsize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE, results="asis"}
model1 <- lm(relig~educ,gss)
model2 <- lm(relig~educ+sex,gss)
model3 <- lm(relig~educ+sex+educ*sex,gss) 
stargazer(model1,model2,model3, out = "text",
          header=FALSE,
          title = "The association between education, sex and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

## Regression

- We can change the dependent variable from religious identification to income

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE, results="asis"}
model1 <- lm(rincome~educ,gss)
model2 <- lm(rincome~educ+sex,gss)
model3 <- lm(rincome~educ+sex+educ*sex,gss) 
stargazer(model1,model2,model3, type = "text",
          header=FALSE,
          title = "The association between education, sex and income",
          digits = 1,
          omit.stat = c("rsq","f","ser"))
```

---

## Regression

\scriptsize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE, results="asis"}
model1 <- lm(rincome~educ,gss)
model2 <- lm(rincome~educ+sex,gss)
model3 <- lm(rincome~educ+sex+educ*sex,gss) 
stargazer(model1,model2,model3, out = "text",
          header=FALSE,
          title = "The association between education, sex and income",
          digits = 1,
          omit.stat = c("rsq","f","ser"))
```


## Plot Fitted Regression Line

- We can use the function `predict.lm` to apply the estimated regression equation to any data
- Here we want to show the predicted values by men and women separately

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
## create a hypothetical data for men
predict_men <-
  data.frame(educ=seq(0,20,1),
           sex=rep(0,21))

## make predictions
y_hat <-
  predict.lm(model3,
           predict_men)

## store the data
predict_men$y_hat <- y_hat
```

---

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
## look at the stored data
head(predict_men)
```

---

- We can symmetrically create predicted values for women

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
## create a hypothetical data for women
predict_women <-
  data.frame(educ=seq(0,20,1),
           sex=rep(1,21))

## make predictions
y_hat <-
  predict.lm(model3,
           predict_women)

## store the data
predict_women$y_hat <- y_hat
```

---

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
## look at the stored data
head(predict_women)
```

---

- We combine these two datasets into one by calling `rbind()` that represents row-wise binding
- To make `rbind()` work, the two datasets have to have same column names

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
predict <- rbind(predict_men,predict_women)
```

---

- Plot the predicted value by calling `ggplot()`

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", warning=FALSE}
library(ggplot2)
ggplot(predict,
       aes(x=educ,y=y_hat,group=sex,color=sex)) +
  geom_line()
  
```

---

- We want to adjust
- 1. `Sex` from a continuous variable to a categorical variable
- 2. Manually defined colors
- 3. `x-axis` and `y-axis` labels
- 4. Omit the legend title
- 5. Change the 

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE}
ggplot(predict,
       aes(x=educ,y=y_hat,group=factor(sex),color=factor(sex))) +
  geom_line() +
  scale_color_manual(labels = c("Men", "Women"), 
                     values = c("darkblue", "darkred")) +
  xlab("Years of education") +
  ylab("Predicted Income") +
  theme_bw() +
  theme(legend.title= element_blank())
```

---

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
ggplot(predict,
       aes(x=educ,y=y_hat,group=factor(sex),color=factor(sex))) +
  geom_line() +
  scale_color_manual(labels = c("Men", "Women"), 
                     values = c("darkblue", "darkred")) +
  xlab("Years of education") +
  ylab("Predicted Income") +
  theme_bw() +
  theme(legend.title= element_blank())
```

---

## Adding more categorical variables

- When we want to add categorical variables that can take multiple values, we use `factor()`
- `R` automatically omits one group as reference

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE}
model4 <- lm(relig~educ+sex+factor(marital),gss)
stargazer(model4, type = "text",
          header=FALSE,
          title = "The association between education, sex, marital status and religion",
          digits = 3,
          single.row = T,
          omit.stat = c("rsq","f","ser"))
```

---

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE, results="asis"}
model4 <- lm(relig~educ+sex+factor(marital),gss)
stargazer(model4, out = "text",
          header=FALSE,
          title = "The association between education, sex, marital status and religion",
          digits = 3,
          single.row = T,
          omit.stat = c("rsq","f","ser"))
```

---

## Plot the Coefficients

- We use function `plot_summs` in `R` to create a coefficient plot
- Take a look at what `plot_summs` can do and its flexibilities [here](https://jtools.jacob-long.com/reference/plot_summs.html)

\footnotesize

```{r,message=FALSE, fig.height=3.5, fig.width=5, fig.align="center", eval=FALSE, warning=FALSE}
library(jtools)
plot_summs(model4,
           colors="darkred",
           coefs = c("Education" = "educ", "Women" = "sex",
                     "Widowed" = "factor(marital)2", 
                     "Divorced" = "factor(marital)3",
                     "Separated" = "factor(marital)4",
                     "Never married" = "factor(marital)5"))
```

---

```{r,message=FALSE, fig.height=3.5, fig.width=5, fig.align="center", echo=FALSE, warning=FALSE}
library(jtools)
plot_summs(model4,
           colors="darkred",
           coefs = c("Education" = "educ", "Women" = "sex",
                     "Widowed" = "factor(marital)2", 
                     "Divorced" = "factor(marital)3",
                     "Separated" = "factor(marital)4",
                     "Never married" = "factor(marital)5"))
```

