---
title: "Week 13: Categorical Data III"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: December 2, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

# Categorical Data as Dependent Variable

## Categorical Data as Independent Variable - Basics

- We talked about the case where categorical data are independent variables two weeks ago
- Instead of (incorrectly) assuming an additive model for categorical data (e.g. race) in a regression model e.g., $Abscale_i = \hat{\beta}_0 + \hat{\beta}_1edu_i + \hat{\beta}_2women_i + \hat{\beta}_3race_i + e_i$, we transform race into 5 "dummy variables" (dummy means a binary variable created from categorical variables), `white`, `black`, `hispanic`, `asian`, `others`, and omit one dummy group as the reference group

---

## Categorical Data as Dependent Variable - Basics

- In many cases, categorical data are dependent variables, such as a binary categorical variable indicating whether the respondent supports legal abortion or has any religious identification \pause
- We may be interested in the association between years of education and religious identification (1=having some identification; 0=no religious identification)
- It is however incorrect to use the OLS model $relig_i = \hat{\beta_0} + \hat{\beta_1}edu_i + e_i$ to estimate the association
- Why?

---

## Categorical Data as Dependent Variable - Basics

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
## set your working directory - you should set your own unique one!
setwd("~/Dropbox/Teaching/SOCUA-302/Week 8")

## read csv data - this is 2021 GSS data
gss <- read.csv("GSS_SOCUA_W8.csv")

## recode religion
gss[which(gss$relig!=4),
    "relig"] <- 1
gss[which(gss$relig==4),
    "relig"] <- 0

gss <- gss %>%
  dplyr::mutate(relig=case_when(educ>16 ~ 0,
                         TRUE ~ relig)) %>%
  filter(educ <= 18) 

gss$random <- rbinom(dim(gss)[1],1,0.1)

gss[which(gss$educ>16&gss$random==1),"relig"] <- 1

gss %>%
  ggplot(aes(x=educ,y=relig)) +
  geom_jitter(width = 0.35,
              height = 0.05,
              shape = 16,
              alpha = 0.5,
              size = 0.1,
              color = "darkblue") +
  geom_smooth(method = "lm",lwd=0.3, se=T,
              color="darkred") +
  theme_bw() +
  xlab("Years of Education") +
  ylab("Having Religious identification") +
  theme(text=element_text(family="Times"),
              plot.title = element_text(size=6),
              axis.text.x=element_text(size=6),
              axis.text.y=element_text(size=6))
  
```

---

## Categorical Data as Dependent Variable - Motivation

- When we treat religious identification, a categorical variable, as a continuous variable and apply an OLS model in estimation, we may have fitted values that do not make sense in reality (e.g., $\hat{y}>1$, or $\hat{y}<0$)
- We want predictions/fitted values to be bounded within 0 and 1, i.e., for a given set of attributes (education), the probability for the respondent to have some religious identification is within 0 and 1 \pause
- This desired property of "boundedness" motivates **logistic** models and **logit** transformations
- (From Lecture) The assumptions of OLS requires that the error terms are normally distributed with a mean of zero and a constant variance, which is impossible when the dependent variable is e.g. dichotomous

---

## Categorical Data as Dependent Variable - Logit Transformation

- Before any transformation, we will need to first think of the predicted/fitted dependent variable as the **probabilities** of having religious identification ($P_{relig}$), given some observed characteristic such as education
  - This is analogous to OLS, where the estimated linear line models the predicted "trend"/"value" of some dependent variable \pause
- Logit transformation: $\log(\frac{P_{relig}}{1-P_{relig}}) = \hat{\beta_0} + \hat{\beta_1}edu_i$ \pause
- But why in this specific form?

---

## Categorical Data as Dependent Variable - Sigmoid Function

- The logit transformation originates from the Sigmoid Function
- $f(x) = \frac{1}{1+e^{-x}}$

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
dat <-
  data.frame(
  x = seq(-10,10,0.05)
)
dat$y <- (1)/(1+exp(-dat$x))
dat %>%
  ggplot(aes(x=x,y=y)) +
  geom_line(color="darkred") +
  theme_bw() +
  xlab("x") +
  ylab("f(x)") +
  ggtitle("Sigmoid Function") +
  theme(text=element_text(family="Times"),
              plot.title = element_text(size=10),
              axis.text.x=element_text(size=6),
              axis.text.y=element_text(size=6))
```

---

## Categorical Data as Dependent Variable - Sigmoid Function

- The logit transformation originates from the Sigmoid Function
- $f(x) = \frac{1}{1+e^{-x}}$
- The Sigmoid Function has two desired properties
  - 1. $f(x)$ is bounded within 0 and 1
  - 2. $x$ has no limit \pause
- The Sigmoid Function is therefore a good candidate to model the **probabilities** of some categorical dependent variable (e.g., having religious identification $P_{relig}$), given some observed characteristic such as education \pause
- We borrow the idea from OLS and estimate $P_{relig_i} = \frac{1}{1+e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}}$ \pause
- With some algebra
- $1-P_{relig_i} = 1-\frac{1}{1+e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}} = \frac{e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}}{1+e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}}$
- $\frac{P_{relig_i}}{1-P_{relig_i}} = \frac{1}{e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}}$
- $\log(\frac{P_{relig_i}}{1-P_{relig_i}}) = \log(\frac{1}{e^{-(\hat{\beta_0} + \hat{\beta_1}edu_i)}}) = \hat{\beta_0} + \hat{\beta_1}edu_i$ \pause
- This is the logit transformation!

---

## Categorical Data as Dependent Variable - Sigmoid and Logit

- Indeed, Sigmoid function and logit function are inverse functions for each other
- Sigmoid function: $y = \frac{1}{1+e^{-x}}$
- Inverse of Sigmoid function: $x = \frac{1}{1+e^{-y}} \rightarrow y = \log(\frac{x}{1-x})$
- $\frac{1}{1+e^{-x}}$ is bounded within 0 and 1. Inversely, the $x$ in $\log(\frac{x}{1-x})$ is bounded within 0 and 1
- Note that the boundedness corresponds to $P_{relig_i}$. $\log(\frac{P_{relig_i}}{1-P_{relig_i}})$ is unbounded

---

## Categorical Data as Dependent Variable - Logit

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
dat <-
  data.frame(
  x = seq(0.0001,0.9999,0.0001)
)
dat$y <- log(dat$x/(1-dat$x))
dat %>%
  ggplot(aes(x=x,y=y)) +
  geom_line(color="darkred") +
  theme_bw() +
  xlab("x") +
  ylab("logit(x)") +
  ggtitle("Logit Function") +
  theme(text=element_text(family="Times"),
              plot.title = element_text(size=10),
              axis.text.x=element_text(size=6),
              axis.text.y=element_text(size=6))
```

---

## Categorical Data as Dependent Variable - Odds and Odds Ratio

- We call the term $\frac{P_{relig_i}}{1-P_{relig_i}}$ in the $\log()$ function “odds” (probability of "event" divided by probability of no "event")
- Odds $\frac{P_{relig_i}}{1-P_{relig_i}} = exp(\hat{\beta_0} + \hat{\beta_1}edu_i)$ \pause
- Odds ratio describes the (multiplicative) change of the odds when the independent variable of interest changes by 1 unit (with other independent variables remain constant in the case of multivariate regression)
- Odds ratio = $\frac{Odds_{edu_i+1}}{Odds_{edu_i}} = \frac{exp(\hat{\beta_0} + \hat{\beta_1}(edu_i+1))}{exp(\hat{\beta_0} + \hat{\beta_1}edu_i)} = exp(\hat{\beta_1})$
- You may find it analogous to OLS, where $\hat{\beta_1}$ describes the (additive) change of the dependent variable when the independent variable change by 1 unit

---

## Categorical Data as Dependent Variable - Interpret Results

- We estimate the logistic regression model 
- $\log(\frac{P_{relig_i}}{1-P_{relig_i}}) = \hat{\beta_0} + \hat{\beta_1}edu_i$

\footnotesize

```{r,message=FALSE, results="asis", echo=FALSE, warning=FALSE}
logit <- glm(relig ~ educ, data = gss, family = "binomial")
library(stargazer)
stargazer(logit, out = "text",
          single.row = F,
          header=FALSE,
          title = "The association between education and religious identification",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

---

- $\log(\frac{P_{relig_i}}{1-P_{relig_i}}) = \hat{\beta_0} + \hat{\beta_1}edu_i = 4.94 - 0.31\times educ$
- A year increase in education is associated with 0.31-unit decrease in $\log(\frac{P_{relig_i}}{1-P_{relig_i}})$
- The associated change of $P_{relig_i}$ is non-linear

---

## Categorical Data as Dependent Variable - Predicted Probabilities

- After logit transformation $\frac{P_{relig_i}}{1-P_{relig_i}} = exp(\hat{\beta_0} + \hat{\beta_1}edu_i)$, the predicted $P_{relig}$ from Maximum Likelihood Estimation relative to years of education becomes

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
logit <- glm(relig ~ educ, data = gss, family = "binomial")
predict <- data.frame(educ=seq(0,18,0.1))
predict$predict_p <- 1/(1+exp(-(summary(logit)$coef[1,1]+summary(logit)$coef[2,1]*predict$educ)))

gss %>%
  ggplot(aes(x=educ,y=relig)) +
  geom_jitter(width = 0.35,
              height = 0.05,
              shape = 16,
              alpha = 0.5,
              size = 0.1,
              color = "darkblue") +
  geom_line(data=predict,aes(x=educ,y=predict_p),
              color="darkred") +
  theme_bw() +
  xlab("Years of Education") +
  ylab("Having Religious identification") +
  theme(text=element_text(family="Times"),
              plot.title = element_text(size=6),
              axis.text.x=element_text(size=6),
              axis.text.y=element_text(size=6))
```

---

## Categorical Data as Dependent Variable - R Operations

- $\log(\frac{P_{relig_i}}{1-P_{relig_i}}) = \hat{\beta_0} + \hat{\beta_1}edu_i$

\footnotesize

```{r,message=FALSE, results="asis", eval=FALSE, warning=FALSE}
logit <- glm(relig ~ educ, data = gss, family = "binomial")
library(stargazer)
stargazer(logit, type = "text",
          single.row = F,
          header=FALSE,
          title = "The association between education and religious identification",
          digits = 3)
```

# Quiz 4 Reviews 

## Exercise for Quiz 4

- How to plot predicted values from OLS when the dependent variable is categorical?
  - Note that in formal statistics, we do not use OLS when estimating categorical dependent variable. We use it here for simplicity
  
---

## Exercise for Quiz 4

- How to calculate Odds, Logit, and probability?
