---
title: "Review for Quiz II"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: October 13, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

# Point Estimate

## Parameters and Estimates

- A **population parameter** describes characteristics of the population ($p$ or $\mu$), which we never observe.
- We use a point **estimate** ($\hat{p}$ or $\hat{\mu}$). calculate using the sample, to estimate the population parameter.

---

## Estimates Uncertainty

- The point **estimate** always includes uncertainty.
- When we draw random samples from the population for a large number of times, the histogram of the sample estimates (e.g., $\hat{p}$s) will be normally distributed. The histogram is the **sampling distribution**. \pause

---

## Estimates Uncertainty

```{r,message=FALSE,echo=FALSE}
## create population, 1 means support, 0 means oppose to legal abortion
population <- c(rep(0,4800000),rep(1,5200000))

## empty vector to store means of sample (K times)
mean_of_sample <- c()

## iterate the process of 3000 times
for (i in 1:3000){
  ## sample with n=1000
  sample <- sample(population,1000)
  mean <- mean(sample)
  mean_of_sample <- c(mean_of_sample,mean)
}
```

```{r,message=FALSE,fig.height=3, fig.width=3, fig.align="center",echo=FALSE}

## plot histogram
hist(mean_of_sample,breaks=20,
     main="Distribution of Sample Proportions",
     xlab="Sample Proportion",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

## Exercise

- The statement is True or False?
  - In real world, we can normally observe the sampling distribution and see if it's a normal distribution.

---

## Estimates Uncertainty

- The normal distribution of the sample proportions (or other sample statistics) are characterized by two important features
  - 1. The sampling distribution is centered around the **population parameter**

```{r,message=FALSE,fig.height=2.8, fig.width=2.8, fig.align="center",echo=FALSE}
## plot histogram
hist(mean_of_sample,breaks=20, main="Distribution of Sample Proportions", xlab="Sample Proportion",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
abline(v=mean(mean_of_sample),col="blue",lwd=2)
```

---

## Estimates Uncertainty

- The normal distribution of the sample proportions (or other sample statistics) are characterized by two important features
  - 2. The standard deviation of the sample proportions depends on the population parameter $p$ and sample size $n$ \pause
  - $SE_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$ \pause
  - We also call it **standard error** of $\hat{p}$ \pause
  - We do not observe $p$, so we use $\hat{p}$ to estimate $SE_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$

---

## Exercise

- The statement is True or False?
  - As the number of observations in a sample increases, the distribution of its values approaches normality. \pause
  - As the number of observations in a sample increases, the (sampling) distribution of its sample means approaches normality. 

---

## Exercise

- In the US, a non-trivial proportion of individuals believe in god. Suppose there are only two types of people in the US population---theists and atheists. A social scientist wants to estimate the proportion of theists in the country. To do so, she samples 1225 individuals randomly from the population, and find that 639 believe in god.
- What is the point estimate?
  - 639/1225 =  0.52
- What is the SE of the point estimate? \pause
  - $\sqrt{\frac{0.52*(1-0.52)}{1225}} = 0.014$

---

## Estimates Uncertainty

- In the normal distribution of sample means, $95\%$ of the $\hat{p}$s will fall into

```{r,message=FALSE,fig.height=3, fig.width=3, fig.align="center",echo=FALSE}
## plot histogram
hist(mean_of_sample,breaks=20, main="Distribution of Sample Proportions", xlab="Sample Proportion",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
abline(v=quantile(mean_of_sample,0.025),col="blue",lwd=2)
abline(v=quantile(mean_of_sample,0.975),col="blue",lwd=2)
```

--- 

## Estimates Uncertainty

- In the normal distribution of sample proportions, $95\%$ of the $\hat{p}$s will fall into
- $p - 1.96 \times SE_{\hat{p}} \le \hat{p} \le p + 1.96 \times SE_{\hat{p}}$ \pause
- Equivalently, $\hat{p} - 1.96 \times SE_{\hat{p}} \le p \le \hat{p} + 1.96 \times SE_{\hat{p}}$ \pause
- This is the $95\%$ **Confidence Interval** for the population $p$

--- 

## Estimates Uncertainty

- We can adjust the $95\%$ confidence interval to e.g., $99\%$ confidence interval. Now, we want an interval where $99\%$ of the $\hat{p}$s fall into

```{r,message=FALSE,fig.height=3, fig.width=3, fig.align="center",echo=FALSE}
## plot histogram
hist(mean_of_sample,breaks=20, main="Distribution of Sample Proportions", xlab="Sample Proportion",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
abline(v=quantile(mean_of_sample,0.025),col="blue",lwd=1)
abline(v=quantile(mean_of_sample,0.975),col="blue",lwd=1)
abline(v=quantile(mean_of_sample,0.005),col="red",lwd=2)
abline(v=quantile(mean_of_sample,0.995),col="red",lwd=2)
```

---

## Estimates Uncertainty

- Looking at the z-score table
- $p - 2.575 \times SE_{\hat{p}} \le \hat{p} \le p + 2.575 \times SE_{\hat{p}}$ \pause
- Equivalently, $\hat{p} - 2.575 \times SE_{\hat{p}} \le p \le \hat{p} + 2.575 \times SE_{\hat{p}}$ \pause
- This is the $99\%$ **Confidence Interval** for the population $p$ 

--- 

## Estimates Uncertainty

- We call $1.96 \times SE_{\hat{p}}$ or $2.575 \times SE_{\hat{p}}$ the **Margin of Error** \pause
- The statement is True or False?
  - At the same level of confidence (e.g., $95\%$), the width of the Confidence Interval is equal to twice of the Margin of Error 
  
---

## Exercise

- A researcher estimated a population parameter $p$ by two samples, with sample size $n_1$ and $n_2$. She constructed $95\%$ Confidence Interval for sample 1, and $99\%$ CI for sample 2. She found that the width of the two CIs are the same.
- What is the relation between $n_1$ and $n_2$?
  - A. $n_1 > n_2$
  - B. $n_1 = n_2$
  - C. $n_1 < n_2$
  - D. Insufficient information

---

## Estimates Uncertainty

- Note that the special formula $SE_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$ only applies to the point estimate of **proportion**.
When the point estimate is about mean (e.g., mean income of a sample),
  - $SE_{\hat{\mu}} = \sqrt{\frac{var(Y)}{n}}$, where $var(Y)$ represent the variance of the population \pause
  - We do not observe $var(Y)$, so we use sample variance $\hat{var}(Y)$ to estimate $SE_{\hat{\mu}} = \sqrt{\frac{\hat{var}(Y)}{n}}$

---

## Exercise

- A fisherman wants to know the mean weight of fish in his fish pond. Knowing that he cannot drain the pond and weigh all fish in a single time, he randomly catches 63 fish, finding that the mean weight of the fish he caught is 21 lbs, and the standard deviation is 2.6 lbs. 
- What is the point estimate? \pause
- What is the SE of the point estimate?
  - $SE_{\hat{\mu}} = \sqrt{\frac{2.6^2}{63}} = 0.33$
  
# Difference b/w two Samples

---

## Difference between two proportions

- Suppose the proportion of men in the population supporting abortion is $p_m$, the proportion of women in the population supporting abortion is $p_w$, and the difference is $p_m-p_w$ \pause
- We draw a random sample from the population with $n_m$ men and $n_w$ women ($n_m+n_w$ in total). 
- We get point estimates $\hat{p}_m$, $\hat{p}_w$, and $\hat{p}_m - \hat{p}_w$ \pause
- What is the standard error of $\hat{p}_m - \hat{p}_w$?
  - $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{p_m(1-p_m)}{n_m} + \frac{p_w(1-p_w)}{n_w}}$, and we estimate it using $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{\hat{p}_m(1-\hat{p}_m)}{n_m} + \frac{\hat{p}_w(1-\hat{p}_w)}{n_w}}$
  
---

## Exercise

- The statement is True or False?
- When we sample from the population for many times and plot the histograms of $\hat{p}_m$, $\hat{p}_w$, and $\hat{p}_m - \hat{p}_w$ calculated from each sample, we will find that $\hat{p}_m$ and $\hat{p}_w$ form bell curves (normal distribution), but $\hat{p}_m - \hat{p}_w$ does not.

---

## Exercise

- To study the gender difference in parental leave, a sociologist surveys 128 men and 254 women with children, finding that $11\%$ men have taken paternity leave, while $65\%$ women have.
- What is the point estimate of the gender difference in parental leave? (i.e., what is the proportion difference?) \pause
  - $0.65 - 0.11 = 0.54$ \pause
- What is the SE of the point estimate? \pause
  - $\sqrt{\frac{0.11*(1-0.11)}{128}+\frac{0.65*(1-0.65)}{254}} = 0.04$

---

## Difference between two Sample Proportions - A Special Case

- What is the standard error of $\hat{p}_m - \hat{p}_w$ **under the null hypothesis** $H_0: p_m = p_w = p$?
- In this **special case** of null hypothesis $H_0: p_m = p_w = p$, we will replace $p_m$ and $p_w$ by a uniform $p$.
- $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{p(1-p)}{n_m} + \frac{p(1-p)}{n_w}}$ \pause
- We do not observe $p$. We estimate $p$ using the sample we drew. 
- We denote this $\hat{p}$ in the case as $\hat{p}_{pooled}$, and $\hat{p}_{pooled} = \frac{\hat{p}_m \times n_m + \hat{p}_w \times n_w}{n_m + n_w}$ \pause
- $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_m} + \frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_w}}$

---

## Difference between two Sample Means

- In many cases, instead of proportion, we are interested in the mean values of two populations and their difference. \pause
- For example, $\mu_1$ could be mean hours worked among men in household production and $\mu_2$ could be mean hours worked among women. We want to know $\mu_1 - \mu_2$ \pause
- We never observe it at the population level. Instead, we gauge it using a male sample and a female sample. The point estimate for the difference of the two sample means is $\bar{Y_1} - \bar{Y_2}$

---

## Difference between two Sample Means

- The standard error of the point estimate $\bar{Y_1} - \bar{Y_2}$, $SE_{\bar{Y_1} - \bar{Y_2}} = \sqrt{\frac{(\sigma_1)^2}{n_1} + \frac{(\sigma_2)^2}{n_2}}$
- $\sigma_1$ is the standard deviation of population 1 (e.g., standard deviation of hours worked of the male population), and $\sigma_2$ is the standard deviation of population 2 (e.g., standard deviation of hours worked of the female population) \pause
- We never observe $\sigma_1$ and $\sigma_2$ \pause
- We estimate it by $SE_{\bar{Y_1} - \bar{Y_2}} = \sqrt{\frac{(s_1)^2}{n_1} + \frac{(s_2)^2}{n_2}}$
- $s_1$ is the standard deviation of sample 1 (e.g., standard deviation of hours worked of the male sample), and $s_2$ is the standard deviation of sample 2 (e.g., standard deviation of hours worked of the female sample)

---

## Exercise

| Gender      | Sample size | Mean of hours | Std Dev |
|-------------|-------------|---------------|---------|
| Men         | 250         | 1.8           | 0.2     |
| Women       | 300         | 4.5           | 0.3     |

- What is the point estimate of the gender difference in mean of hours? \pause
  - $\bar{Y}_1-\bar{Y}_2=4.5-1.8=2.7$ \pause
- What is $SE_{\bar{Y}_1-\bar{Y}_2}$? \pause
  - $SE_{\bar{Y}_1-\bar{Y}_2} = \sqrt{\frac{(s_1)^2}{n_1} + \frac{(s_2)^2}{n_2}} = \sqrt{\frac{0.2^2}{250}+\frac{0.3^2}{300}} = 0.02$ \pause
- What is the $95\%$ Confidence Interval of $Y_1-Y_2$? \pause
  - $\bar{Y}_1-\bar{Y}_2 - 1.96 \times SE_{\bar{Y}_1-\bar{Y}_2} \le Y_1-Y_2 \le \bar{Y}_1-\bar{Y}_2 + 1.96 \times SE_{\bar{Y}_1-\bar{Y}_2}$, i.e., $2.66 \le Y_1-Y_2 \le 2.74$

# Hypothesis Testing

---

## Terminology

- The **null hypothesis** $H_0$ often states "nothing is going on here"
- The **alternative hypothesis** $H_A$ represents what we really think is going on---a substantive alternative to the null \pause
- For example, we want to know the gender difference in the hours devoted to housework
  - $H_0$: there is no gender difference **in the population**
  - $H_A$: there is gender difference **in the population**
  
---

## Terminology

- We use a sample to estimate a population property, and decide whether to reject $H_0$ or not.
- Point estimate always involves uncertainties, so one may falsely reject the null hypothesis or fail to reject the null hypothesis

| Truth         | Do not reject $H_0$ | Reject $H_0$ in favor of $H_A$  |
|---------------|---------------------|---------------------------------|
| $H_0$ is true | Right decision      | Type 1 error                    |
| $H_A$ is true | Type 2 error        | Right decision                  | 

---

## Hypothesis Testing

- Hypothesis testing focuses on $H_0$ and the probability of **Type 1 error** \pause
- In a more plain language, it means that, **given that the null hypothesis is true**, what is the probability of getting the test statistics.
  - E.g., if $H_0$: there is no gender differences in yearly income, yet we observe from our sample that $\bar{Y}_1 - \bar{Y}_2 = 10,000$, it is reasonable to speculate that $H_0$ is incorrect (i.e., we reject $H_0$ and accept $H_A$: there are gender differences in yearly income at the level of population). \pause
  - The lower the probability, the less likely the null hypothesis is true.
  - The higher the probability, the more likely the null hypothesis is true. \pause
- How low should this probability to be, such that we are "confident" enough to reject the null hypothesis? \pause
  - This is called **Significance level** $(\alpha)$, usually pre-set at 0.05 ($5\%$). When the probability is lower than 0.05, we are confident to reject the null hypothesis.
  
---

## Hypothesis Testing

- What does it mean to say, when the probability is lower than 0.05, we are confident to reject the null hypothesis? \pause
- We sample many times ($K$ times) from the population, assuming that there is no gender difference in yearly income. In $95\%$ of the $K$ point estimates, $\bar{Y}_1 - \bar{Y}_2$ falls within $[lower \: bar, higher  \: bar]$, e.g., $[-1000,1000]$
- The point estimate of the single sample that we draw, however, reports that $\bar{Y}_1 - \bar{Y}_2 = 10,000$. This is very different from 0 and the times of getting $\bar{Y}_1 - \bar{Y}_2 = 10,000$ are much smaller than $0.05K$. We therefore reject $H_0$ at the Significance level $\alpha=0.05$.

---

## Hypothesis Testing

- How do we calculate the probability of getting the sample statistics, given the null hypothesis?
- This is where $z$-score, the $z$-score table, and $p$-value step in \pause
- Suppose we are interested in the gender difference in yearly income. We sample 1200 men and 1000 women, finding that $\bar{Y}_{men} = 50000$ and $\bar{Y}_{women} = 49000$. We also find that $s_{men} = 8000$ and $s_{women}=7000$ \pause
  - 0. Calculate the point estimate $\bar{Y}_{men} - \bar{Y}_{women} = 50000-49000=1000$, and $SE_{\bar{Y}_{men}-\bar{Y}_{women}} = \sqrt{\frac{8000^2}{1200} + \frac{7000^2}{1000}} = 319.9$ \pause
  - 1. Look at the null hypothesis. In this case, $H_0$: $\bar{Y}_{men}-\bar{Y}_{women} = 0$. \pause
  - 2. Calculate the $z$-score: $z = \frac{point \: estimate - null\:value}{SE} = \frac{\bar{Y}_{men} - \bar{Y}_{women} - 0}{SE_{\bar{Y}_{men}-\bar{Y}_{women}}} = \frac{1000}{319.9} = 3.13$ \pause
  - 3. Check the $p$-value associated with the $z$-score. We find 0.0009, meaning that if there is no gender difference in the population, the chance that we get the above sample statistics is 0.0009. \pause
  - 4. Note that to get a difference of 1000, it is also possible that women's income lead men's. The possibility to get such a deviation (1000) given the null hypothesis $Y_{men}=Y_{women}$ is therefore $0.0009 \times 2 = 0.0018$. This is the $p$-value = $0.0018$. \pause
  
---

## Hypothesis Testing

- $p$-value = $0.0018$ is smaller than $\alpha=0.05$. We reject the null hypothesis at the 0.05 significance level.
- Equivalently, the $95\%$ confidence interval of $Y_{men} - Y_{women}$ does not contain 1000 under the null hypothesis. 

---

## Exercise

- Different countries in the world have different sex ratios at birth. A scholar wants to know whether country C has a skewed sex ratio at birth or not. He samples 1200 infants from the country and finds that $53\%$ of the infants in the sample are boys and $47\%$ are girls.
- What is the null hypothesis? \pause
- What is the point estimate of the boy proportion? \pause
- What is the SE of the point estimate? \pause
- What is the $z$-score of the point estimate under the null hypothesis? \pause
- What is the $p$-value associated with the $z$-score? \pause
- Can we reject the null hypothesis at $\alpha=0.01$? \pause
- Does the $95\%$ confidence interval of $p$ under the null hypothesis include $\hat{p}$? What about $99\%$ CI?
  

