---
title: "Week 3: Summarizing Data II"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: September 23, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

# Data 

## Data Structures

- `R` also has a number of basic data *structures*. 
- A data structure is either 
    - homogeneous (all elements are of the same data type) 
    - heterogeneous (elements can be of more than one data type).

| Dimension | **Homogeneous** | **Heterogeneous** |
|-----------|-----------------|-------------------|
| 1         | Vector          | List              |
| 2         | Matrix          | Data Frame        |
| 3+        | Array           |                   |


# Summarize Data (Basics)

## Read Data

- A **data frame** is the **most** common way that we store and interact with data

```{r, message = FALSE, warning = FALSE}
## set working directory
setwd("~/Dropbox/Teaching/SOCUA-302/Week 2")

## read the file
gss <- read.csv("GSS_SOCUA_W2.csv")
```

---

## Combine data subsetting and summarizing (base R)

- Some values in the variable `sibs` are not real observations
- Negative values are missingness
- How do we calculate the mean value of `sibs` without negative `sibs` values?

```{r}
mean(gss[gss$sibs>=0,"sibs"])
```

---

## Exercise

- How do we calculate the variance of `sibs` in year 2018 only?

---

## Combine data subsetting and summarizing (dplyr)

- We can also use `dplyr` to subset and summarize data
  - Remember `filter()` filters rows and `select()` select columns
- How do we calculate the mean value of `sibs` without negative `sibs` values?

```{r,message=FALSE}
library(dplyr)
gss %>% filter(sibs>0) %>% summarize(sibs_mean = mean(sibs))
```

- `summarize()` follows the format: `summarize(your_summarize_name = function(variable))`

---

## Combine data subsetting and summarizing (dplyr)

- How do we calculate the mean value of `sibs` without negative `sibs` values?
- Why bothering using `dplyr` and `summarize`?
  - Because we can summarize data using different functions at the same time

```{r,message=FALSE}
gss %>% filter(sibs>0) %>% summarize(sibs_mean = mean(sibs),
                                     sibs_var = var(sibs))
```

---

## Export the data

- `write.csv` save (or export) the dataframe in `.csv` format.

```{r, eval=FALSE}
write.csv(gss, "cleaned_gss.csv")
```

---

## Histogram plot

- Histograms describe the distribution of **numeric** data

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss$age)
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss$age, cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- Again, we want to drop negative values of `age`
- We may also want to see the distribution for some years but not others

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"])
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- To change plot titles and x-axis label

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", 
     xlab = "Age")
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", 
     xlab = "Age",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- To control the number of `bins`

\small

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS",
     xlab = "Age",
     breaks = 30)
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", xlab = "Age",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5,
     breaks = 30)
```

---

## Barplot

- Barplots describe the distribution of **categorical** data

```{r, fig.height=3, fig.width=3, fig.align="center"}
## create a count summary in each category by function `table`
counts <- table(gss[gss$age>0 & gss$year<=2018 &
           gss$year>=2012 & gss$marital>=0, "marital"])
counts
```

---

\small

```{r, fig.height=3, fig.width=3, fig.align="center"}
barplot(counts, main="Distribution of Marital Status in 2012-2018, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Widowed", "Divorced", "Separated", 
               "Never Married"),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.32)
```

--- 

- We can also change the order of the categories by first `factorizing` the variable with manual `levels`

```{r}
gss$marital <- factor(gss$marital,levels = c(1,3,4,2,5))
counts <- table(gss[gss$age>0 & gss$year<=2018 &
           gss$year>=2012, "marital"])
```

---

\small

```{r, fig.height=3, fig.width=3, fig.align="center"}
barplot(counts, main="Distribution of Marital Status in 2012-2018, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Divorced", "Separated", "Widowed", "Never Married"),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.35)
```

# Summarize Data (Advanced)

---

## Summarize data by group

- In many cases, we not only want data summaries for the total sample, but summaries for some sub-samples
- For example, we may want to summarize religious preference by gender (the column `relig`)
- In the `relig` column, $4$ means no religious belief, and other non-zero numbers mean some religious belief
- It is a non-ordered categorical variable

```{r}
unique(gss$relig)
```

## Summarize data by group

- As there are many religious categories, we may only want two groups categorization for now: religious and non-religious individuals

```{r}
gss[gss$relig!=4 & gss$relig>0,
    "relig"] <- 1
gss[gss$relig==4,
    "relig"] <- 0
unique(gss$relig)
```

---

## Summarize data by group

- We will use the `group_by()` function in `dplyr` to summarize data by the group we specify
- It can be nested in a `%>%` pipeline

```{r,message=FALSE}
gss %>%
  filter(relig>=0&sex>0) %>%
  group_by(sex) %>%
  summarize(religious = mean(relig))
```

---

## Exercise

- The calculation above includes all individuals surveyed since 1972 until 2021
- How to get the same statistic, but only in 2021?


---

## Exercise

- The variable `relig16` asks which religion was the respondent raised in
- How to get the proportion of religious people (at the time of being interviewed, the column `relig`) by whether they were raised in a religious family? 
- Do not forget to drop missing values first!

---

## Exercise

```{r}
gss[gss$relig16!=4&gss$relig16>=0,"relig16"] <- 1
gss[gss$relig16==4,"relig16"] <- 0
gss %>%
  filter(relig>=0&sex>0&relig16>=0) %>%
  group_by(relig16) %>%
  summarize(religious = mean(relig))
```

## Exercise

- The variable `relig16` asks which religion was the respondent raised in
- How to get the proportion of religious people (at the time of being interviewed, the column `relig`) by whether they were raised in a religious family?
- How to add another column summarizing the proportion of non-religious people by whether they were raised in a religious family?

---

## Exercise

\small

```{r}
gss[gss$relig16!=4&gss$relig16>=0,"relig16"] <- 1
gss[gss$relig16==4,"relig16"] <- 0
gss %>%
  filter(relig>=0&sex>0&relig16>=0) %>%
  group_by(relig16) %>%
  summarize(religious = mean(relig),
            nonreligious=mean(1-relig))
```

---

## Viasualizing the trend

- We have already calculated the proportion of religious men and women in all and one specific year(s)
- Another common exploration of such data is to analyze the temporal trend of the proportion of religious people by gender over the years

---

```{r,message=FALSE}
gss %>%
  filter(relig>=0&sex>0) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  head()
```

## Viasualizing the trend (ggplot2)

- While base `R` can handle such visualizations, we will introduce another powerful visualization package, `ggplot2`
- `ggplot2` can also be nested in the `%>%` pipeline
- It is highly flexible in changing colors, setting fonts, customing legends, etc.

---

## Viasualizing the trend (ggplot2)

- The basic structure of `ggplot2` is

```{r,eval=FALSE}
## create a ggplot object
ggplot(data,
       aes(x=variable1,
           y=variable2,
           group=variable3,
           ...)) +
  geom_line() +
  geom_point() +
  ...

```

--- 

- The basic **parameters** such as x-axis, y-axis, groups and colors, are controlled by `aes()` that stands for aesthetics
- To create a trend plot, we need a **line** plot controlled by `geom_line()`
- There are many other options, including `geom_point()` that creates scatter plot, `geom_histogram()` that creates histograms, etc.
- Check this website for much more details of what `ggplot2` can do!
- https://ggplot2.tidyverse.org/reference/

---

## Viasualizing the trend (ggplot2)


```{r,message=FALSE, fig.height=3, fig.width=3, fig.align="center", eval=FALSE}
library(ggplot2)
gss %>%
  filter(relig>=0&sex>0) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,group=sex,color=sex)) +
  geom_line()
```

---

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", echo=FALSE}
library(ggplot2)
gss %>%
  filter(relig>=0&sex>0) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,group=sex,color=sex)) +
  geom_line()
```

---

- It looks fine, but need some refinements! \pause
- R identifies the variable `sex` as a continuous variable, we need a categorical variable
  - We can change the data type of `sex` into characters
  - We use function `mutate(new_variable = ...)` in `dplyr` to create new variable(s)
  - The new variable name can be the same as the original one (i.e., overwriting)

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", eval=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  geom_line()
```

---

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", echo=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) + ## note here you do not need group anymore
  geom_line()
```

---

- Readers do not know what 1 and 2 in sex means! We need `male` and `female` labels
- We may also want to change the line color

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", eval=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred"))
  geom_line()
```

---

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", echo=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  geom_line() +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred"))
```

---

- The background looks a bit pale...
- And we need a title

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", eval=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred")) +
  geom_line() +
  theme_bw() +
  ggtitle("The Proportion of Religious People \n
          by Gender, 1972-2021, GSS")
```

---

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", echo=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred")) +
  geom_line() +
  theme_bw() +
  ggtitle("The Proportion of Religious People 
          by Gender, 1972-2021, GSS")
```

---

With more tweaks, we can make the figure publishable

\small

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", eval=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  geom_line() +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred")) +
  theme_bw() + 
  ggtitle("The Proportion of Religious People by Gender, 1972-2021, GSS") +
  xlab("Year") +
  ylab("Proportion with Religious Preferences") +
  theme(plot.title=element_text(size=10),
        text=element_text(family="Times"),
        axis.title.x=element_text(size=8),
        axis.title.y=element_text(size=8),
        legend.title = element_text(size=8),
        legend.text = element_text(size=6))
```

---

```{r,message=FALSE, fig.height=3, fig.width=5, fig.align="center", echo=FALSE}
gss %>%
  filter(relig>=0 & sex>=0) %>%
  mutate(sex = as.character(sex)) %>%
  group_by(year, sex) %>%
  summarize(religious = mean(relig)) %>%
  ggplot(aes(x=year,y=religious,color=sex)) +
  geom_line() +
  scale_color_manual(labels = c("Male", "Female"), 
                     values = c("darkblue", "darkred")) +
  theme_bw() + 
  ggtitle("The Proportion of Religious People by Gender, 1972-2021, GSS") +
  xlab("Year") +
  ylab("Proportion with Religious Preferences") +
  theme(plot.title=element_text(size=10),
        text=element_text(family="Times"),
        axis.title.x=element_text(size=10),
        axis.title.y=element_text(size=10),
        legend.title = element_text(size=10),
        legend.text = element_text(size=8))
```

