---
title: "Week 10: Categorical Data"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: November 11, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

# Quiz 3

## Common Mistakes

- True or False
- A significant regression coefficient for an independent variable $X$ indicates that $X$ is a cause of the dependent variable $y$

---

## Read STATA output

- Look at the STATA output

---

## Intuition of Multivariate Regression

- When the second independent variable is a categorical variable with three possible values; slopes do not differ by the second independent variable

```{r,message=FALSE, fig.height=4, fig.width=5.5, fig.align="center", echo=FALSE, warning=FALSE}
library(scatterplot3d)
## create a population
x1 <- runif(10000, min=0, max=20)
x2 <- rbinom(10000, size=2, prob=0.5)
beta1 <- 4000
beta2 <- 20000
beta0 <- 25000
epsilon <- rnorm(10000,mean=0,sd=10000)
y <- beta0+beta1*x1+beta2*x2+epsilon
y<-y/1000

## draw a sample
set.seed(1)
pop <- data.frame(education=x1,race=x2,income=y)
sample <- pop[sample(nrow(pop), 1000), ]

## plot the sample-level correlation
colors <- c("#999999", "#E69F00","darkblue")
colors <- colors[as.numeric(sample$race)+1]
scatterplot3d(sample, pch = 16, 
              color=colors,
              cex.symbols=0.4*par("cex"),
              cex.axis=0.5*par("cex.axis"),
              cex.lab=0.5*par("cex.lab"), 
              font.axis=0.5*par("font.axis"),
              font.lab=0.5*par("font.lab"))
```

---

## Intuition of Multivariate Regression

- We can visualize the 3-D plot by a 2-D scatterplot

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
## plot the association by race
sample %>%
  ggplot(aes(y=income,x=education,group=factor(race), 
             color=factor(race))) +
  geom_point(size=0.5) +
  scale_color_manual(values=c("#999999", "#E69F00","darkblue"),
                     labels = c("Asian","Black","White"),
                     name = "") +
  geom_smooth(method = "lm",lwd=0.3) +
  theme_bw()
```

---

## Intuition of Multivariate Regression

- Without the hypothetical points, we get similar lines as in the quiz

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
## plot the association by race
sample %>%
  ggplot(aes(y=income,x=education,group=factor(race), 
             color=factor(race))) +
  scale_color_manual(values=c("#999999", "#E69F00","darkblue"),
                     labels = c("Asian","Black","White"),
                     name = "") +
  geom_smooth(method = "lm",lwd=0.3, se=F) +
  theme_bw()
```

---

## Intuition of Multivariate Regression

- When the second independent variable is a categorical variable with three possible values; slopes differ by the second independent variable

```{r,message=FALSE, fig.height=4, fig.width=5.5, fig.align="center", echo=FALSE, warning=FALSE}
library(scatterplot3d)
## create a population
x1 <- runif(10000, min=0, max=20)
x2 <- rbinom(10000, size=2, prob=0.5)
beta1 <- 4000
beta2 <- 20000
beta3 <- -2000
beta0 <- 25000
epsilon <- rnorm(10000,mean=0,sd=10000)
y <- beta0+beta1*x1+beta2*x2+beta3*x1*x2+epsilon
y<-y/1000

## draw a sample
set.seed(1)
pop <- data.frame(education=x1,race=x2,income=y)
sample <- pop[sample(nrow(pop), 1000), ]

## plot the sample-level correlation
colors <- c("#999999", "#E69F00","darkblue")
colors <- colors[as.numeric(sample$race)+1]
scatterplot3d(sample, pch = 16, 
              color=colors,
              cex.symbols=0.4*par("cex"),
              cex.axis=0.5*par("cex.axis"),
              cex.lab=0.5*par("cex.lab"), 
              font.axis=0.5*par("font.axis"),
              font.lab=0.5*par("font.lab"))
```

---

## Intuition of Multivariate Regression

- We can visualize the 3-D plot by a 2-D scatterplot

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
## plot the association by race
sample %>%
  ggplot(aes(y=income,x=education,group=factor(race), 
             color=factor(race))) +
  scale_color_manual(values=c("#999999", "#E69F00","darkblue"),
                     labels = c("Asian","Black","White"),
                     name = "") +
  geom_point(size=0.5) +
  geom_smooth(method = "lm",lwd=0.3, se=T) +
  theme_bw()
```

---

## Intuition of Multivariate Regression

- Without the hypothetical points, we get similar lines as in the quiz

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
## plot the association by race
sample %>%
  ggplot(aes(y=income,x=education,group=factor(race), 
             color=factor(race))) +
  scale_color_manual(values=c("#999999", "#E69F00","darkblue"),
                     labels = c("Asian","Black","White"),
                     name = "") +
  geom_smooth(method = "lm",lwd=0.3, se=F) +
  theme_bw()
```


# Categorical Data

## Basics

- While we already talked much about quantitative/numeric/continuous data (e.g., income), we have not discussed much about categorical data
- In statistics, a categorical variable is a variable that can take on one of a limited, and usually fixed, number of possible values, assigning each unit of observation to a particular group or nominal category on the basis of some qualitative property
  - Typical examples are gender, race, class (working class, middle class, upper class), and religious preferences
  - Sometimes these variables are the core of contemporary sociology \pause
- why do we specifically care about categorical data beyond numeric data?

## Basics

- We may be interested in the association between years of education and religious belief (1=having some belief; 0=no religious belief)

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE}
## set your working directory - you should set your own unique one!
setwd("~/Dropbox/Teaching/SOCUA-302/Week 8")

## read csv data - this is 2021 GSS data
gss <- read.csv("GSS_SOCUA_W8.csv")

## recode religion
gss[which(gss$relig!=4),
    "relig"] <- 1
gss[which(gss$relig==4),
    "relig"] <- 0

gss %>%
  dplyr::mutate(relig=case_when(educ>16 ~ 0,
                         TRUE ~ relig)) %>%
  filter(educ <= 18) %>%
  ggplot(aes(x=educ,y=relig)) +
  geom_jitter(width = 0.35,
              height = 0.05,
              shape = 16,
              alpha = 0.5,
              size = 0.1,
              color = "darkblue") +
  geom_smooth(method = "lm",lwd=0.3, se=T,
              color="darkred") +
  theme_bw() +
  xlab("Years of Education") +
  ylab("Having Religious Belief") +
  theme(text=element_text(family="Times"),
              plot.title = element_text(size=6),
              axis.text.x=element_text(size=6),
              axis.text.y=element_text(size=6))
  
```

---

## Basics

- We may be interested in the association between years of education and religious belief
- When we treat religious belief, a categorical variable, as a continuous variable, we may have fitted values that do not make sense in reality (e.g., $\hat{y}>1$, or $\hat{y}<0$)
- We want predictions to be bounded within 0 and 1
- We will go beyond OLS in the following weeks to address these particular scenarios stemming from categorical data

---

## Basics

- How do we describe categorical data?
- For a single variable
  - For numeric/quantitative data, we use histogram \pause
  - For categorical data, we use **barplot** 
- We use **barplot** to describe the distribution of categorical values (e.g., gender, race, marital status)

---

## Barplot

- Barplots describe the distribution of categorical data

```{r, fig.height=3, fig.width=3, fig.align="center",echo=FALSE}
## create a count summary in each category by function `table`
counts <- table(gss$marital)

## barplot
barplot(counts, main="Distribution of Marital Status in 2021, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Widowed", "Divorced", "Separated", 
               "Never Mar."),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.32)
```

---

## Basics

- How do we describe categorical data?
- For a single variable
  - For numeric/quantitative data, we use histogram \pause
  - For categorical data, we use **barplot** 
- We use **barplot** to describe the distribution of categorical values (e.g., gender, race, marital status)
- For two categorical variables (e.g., gender current religious belief)
  - We use a cross-table to summarize the relationship

---

## Basics

- The proportion of men and women who have and have no religious belief

```{r}
gss %>%
  filter(!is.na(sex)) %>%
  group_by(sex) %>%
  summarize(religious = mean(relig,na.rm=T),
            nonreligious=mean(1-relig,na.rm=T))
```

---

## t-test

- Suppose we want to know the gender difference in religious belief
- This is a typical problem of testing whether **two samples** differ in a proportion \pause
- The point estimate is $\hat{p}_w - \hat{p}_m$
- Under the null hypothesis $H_0: p_m = p_w = p$, the estimated $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_m} + \frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_w}}$, where
- $\hat{p}_{pooled} = \frac{\hat{p}_m \times n_m + \hat{p}_w \times n_w}{n_m + n_w}$

---

## t-test
- Now in 2021 GSS, there are 1736 men and 2204 women
- $\hat{p}_{pooled} = \frac{\hat{p}_m \times n_m + \hat{p}_w \times n_w}{n_m + n_w} = \frac{0.680 \times 1736 + 0.745 \times 2204}{1736 + 2204}=0.716$
- $SE_{\hat{p}_m - \hat{p}_w} = \sqrt{\frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_m} + \frac{\hat{p}_{pooled}(1-\hat{p}_{pooled})}{n_w}} = 0.014$
- The point estimate $\hat{p}_w - \hat{p}_m = 0.745- 0.680= 0.065$ \pause
- $t = \frac{0.065-0}{0.0145} = 4.64$
- Do we reject the null hypothesis?

---

## Regression

- The t-test produces the exact same estimates as OLS regression
- $relig_i = \hat{\beta}_0 + \hat{\beta}_1gender_i +e_i$

\footnotesize

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, warning=FALSE, results="asis"}
library(stargazer)
model <- lm(relig~sex,gss)
stargazer(model, out = "text",
          single.row = F,
          header=FALSE,
          title = "The association between gender and religious belief",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

# R Operations

## Read Data

```{r}
## set your working directory - you should set your own unique one!
setwd("~/Dropbox/Teaching/SOCUA-302/Week 8")

## read csv data - this is 2021 GSS data
gss <- read.csv("GSS_SOCUA_W8.csv")
```

---

## Barplot

```{r, fig.height=3, fig.width=3, fig.align="center",eval=FALSE}
## create a count summary in each category by function `table`
counts <- table(gss$marital)

## barplot
barplot(counts, main="Distribution of Marital Status in 2021, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Widowed", "Divorced", "Separated", 
               "Never Mar."),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.32)
```

---

## Cross Table

- The proportion of men and women who have and have no religious belief

\footnotesize

```{r}
## recode religion
gss[which(gss$relig!=4),"relig"] <- 1
gss[which(gss$relig==4),"relig"] <- 0

## cross table
gss %>%
  filter(!is.na(sex)) %>%
  group_by(sex) %>%
  summarize(religious = mean(relig,na.rm=T),
            nonreligious=mean(1-relig,na.rm=T))
```

---

## Regression

- The t-test produces the exact same estimates as OLS regression
- $relig_i = \hat{\beta}_0 + \hat{\beta}_1gender_i +e_i$

```{r,message=FALSE, fig.height=2.5, fig.width=4, fig.align="center", eval=FALSE, warning=FALSE, results="asis"}
library(stargazer)
model <- lm(relig~sex,gss)
stargazer(model, type = "text",
          single.row = F,
          header=FALSE,
          title = "The association between gender and religious belief",
          digits = 3,
          omit.stat = c("rsq","f","ser"))
```

