---
title: "Week 2: Summarizing Data"
author: "Wenhao Jiang"
institute: |
           | Department of Sociology
           | New York University
date: September 17, 2022
output: beamer_presentation
classoption: "aspectratio=1610"
theme: "default"
outertheme: "miniframes"
header-includes:
  - \definecolor{nagivation}{rgb}{0,0,0.3}
  - \definecolor{main}{rgb}{0,0,0.5}
  - \setbeamercolor{section in head/foot}{bg=nagivation,fg=white}
  - \usecolortheme[named=main]{structure}
  - \setbeamertemplate{footline}[page number]
  - \setbeamerfont{footnote}{size=\footnotesize}
  - \setbeamerfont{caption}{size=\footnotesize}
---

# Data 

## Data Structures

- `R` also has a number of basic data *structures*. 
- A data structure is either 
    - homogeneous (all elements are of the same data type) 
    - heterogeneous (elements can be of more than one data type).

| Dimension | **Homogeneous** | **Heterogeneous** |
|-----------|-----------------|-------------------|
| 1         | Vector          | List              |
| 2         | Matrix          | Data Frame        |
| 3+        | Array           |                   |


# Data frame

## Data frame basics

- A **data frame** is the **most** common way that we store and interact with data

```{r}
example_data <- data.frame(x = c(1, 3, 5, 7, 9, 1, 3, 5, 7, 9),
                          y = c(rep("Hello", 9), "NYU"),
                          z = rep(c(TRUE, FALSE), 5))
```

- A data frame is a collection of vertical vectors
    - Each vector must contain the same data type
    - The difference vectors can store different data types

---

```{r}
example_data
```

---

## Read data frame

- In most cases, we do not deal with self-created data frames
- Instead, we use external data (e.g., surveys) and import them to `R`/`RStudio`

---

## Read data frame

- External data can be stored in different formats
- One typical format is `.csv`
- Read `.csv` file into R
    - `read.csv()` function as default
    - `read_csv()` function from the `readr` package. This is faster for larger data

---

## Read data frame

- But before that, you should let `R` know where the file is located
- In `R`, we call the **folder** that `R` currently is reading **the working directory** 
- To see the current working directory, type in

```{r}
getwd()
```

---

## Read data frame

- If you want to re-set the working directory to the folder that you store the file you want to read, use `setwd()` function

```{r, eval = FALSE}
setwd()
```

- `RStudio` also allows a more "intelligent" way to reset the working directory

```{r, message = FALSE, warning = FALSE}
## set working directory
setwd("~/Dropbox/Teaching/Week 2")

## read the file
gss <- read.csv("GSS_SOCUA_W2.csv")
```

---

## Examine dataframe basics

- To see the first n rows of the data
```{r}
head(gss, n=5)
```

--- 

- To see the last n rows of the data
```{r}
tail(gss, n=5)
```

- To view the whole dataset
```{r,eval=FALSE}
View(gss)
```

- `View()` function generally not recommended, especially if the data is large

--- 

## Examine dataframe basics

- To check the general "structure" of the data frame
```{r}
str(gss)
```

- `str()` will display the number of **observations** and **variables**, list the variables, give the type of each variable, and show some elements of each variable

---

## Examine dataframe basics

-  `colnames()` function to obtain names of the variables in the dataset
```{r}
colnames(gss)
```

- To access one of the variables **as a vector**, we use the `$` operator
```{r,eval=FALSE}
gss$sibs
```

--- 

- We can use `min()` or `max()` to inspect the miminum and maximum value of the variable
```{r}
min(gss$sibs)
max(gss$sibs)
```

- We can use `unique()` to inspect the unique values of the variable
```{r}
unique(gss$marital)
```

- You can quickly spot some unusual observartions. Deal with them carefully!

---

- We can use the `dim()`, `nrow()` and `ncol()` functions to obtain information about the dimension of the data frame

```{r}
dim(gss)
nrow(gss)
ncol(gss)
```

## Subsetting data

- Subsetting data frames can work much like subsetting matrices using square brackets, `[,]`. 
- But instead of using column indexes, we use column names
- We can select a single column by

```{r,eval=FALSE}
gss[gss$year > 2018, "marital"]
```

---

- We can also select multiple columns by creating a vector for column names
```{r,eval=FALSE}
gss[gss$year > 2018, c("marital","year")]
```

---

## Package dplyr

- To subset data, another approach specifically designed for data frames is calling the `filter` and `select` functions from the `dplyr` package
- `select` function selects certain **columns**
- `filter` function filters certain **rows** (or **observations**) based on some conditions

```{r, eval = FALSE}
library(dplyr)
gss %>% 
  filter(year>2018) %>% 
  select(marital,sibs,age,sex)
```

# Summarize Data

---

## Mean, Median, Variance

- To calculate mean, median, and variance, `R` uses functions

```{r}
mean(gss$sibs)
median(gss$sibs)
var(gss$sibs)
```

---

- Note that the denominator of variance is `n-1` by default in `R`
$var(x) = \frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2$

```{r}
var(c(1,2,3))
```

---

## Combine data subsetting and summarizing (base R)

- Some values in the variable `sibs` are not real observations
- Negative values are missingness
- How do we calculate the mean value of `sibs` without negative `sibs` values? \pause

```{r}
mean(gss[gss$sibs>=0,"sibs"])
```

---

## Exercise

- How do we calculate the variance of `sibs` in year 2018 only?

---

## Combine data subsetting and summarizing (dplyr)

- We can also use `dplyr` to subset and summarize data
- How do we calculate the mean value of `sibs` without negative `sibs` values?

```{r,message=FALSE}
library(dplyr)
gss %>% filter(sibs>0) %>% summarize(sibs_mean = mean(sibs))
```

- `summarize()` follows the format: `summarize(your_summarize_name = function(variable))`

---

## Histogram plot

- Histograms describe the distribution of **numeric** data

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss$age)
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss$age, cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- Again, we want to drop negative values of `age`
- We may also want to see the distribution for some years but not others

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"])
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- To change plot titles and x-axis label

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", 
     xlab = "Age")
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", 
     xlab = "Age",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5)
```

---

- To control the number of `bins`

\small

```{r, fig.height=3, fig.width=3, eval=FALSE}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS",
     xlab = "Age",
     breaks = 30)
```

```{r, fig.height=3, fig.width=3, echo=FALSE, fig.align="center"}
hist(gss[gss$age>0 & gss$year<=2018 & gss$year>=2012,"age"],
     main = "Distribution of Respondents Ages in 2012-2018, GSS", xlab = "Age",
     cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5,
     breaks = 30)
```

---

## Barplot

- Barplots describe the distribution of **categorical** data

```{r, fig.height=3, fig.width=3, fig.align="center"}
## create a count summary in each category by function `table`
counts <- table(gss[gss$age>0 & gss$year<=2018 &
           gss$year>=2012 & gss$marital>=0, "marital"])
counts
```

---

\small

```{r, fig.height=3, fig.width=3, fig.align="center"}
barplot(counts, main="Distribution of Marital Status in 2012-2018, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Widowed", "Divorced", "Separated", 
               "Never Married"),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.32)
```

--- 

- We can also change the order of the categories by first `factorizing` the variable with manual `levels`

```{r}
gss$marital <- factor(gss$marital,levels = c(1,3,4,2,5))
counts <- table(gss[gss$age>0 & gss$year<=2018 &
           gss$year>=2012, "marital"])
```

---

\small

```{r, fig.height=3, fig.width=3, fig.align="center"}
barplot(counts, main="Distribution of Marital Status in 2012-2018, GSS",
   xlab="Marital Status",
   names.arg=c("Married", "Divorced", "Separated", "Widowed", "Never Married"),
   cex.lab=0.5, cex.axis=0.5, cex.main=0.5, cex.sub=0.5, cex.names=0.35)
```

---

## Export the data

- `write.csv` save (or export) the dataframe in `.csv` format.

```{r, eval=FALSE}
write.csv(gss, "cleaned_gss.csv")
```

---

- We will cover more advanced data summarization next week, including how we can summarize data by group
- We will also cover package `ggplot2` next week, a powerful and more flexible tool in data visualization, including what we introduced today

---

- Any questions from textbook or lectures?

